DOpElib: Differential Equations and Optimization Environment; A Goal Oriented Software Library for Solving PDEs and Optimization Problems

### 2012-11-27 Paper_May_2012/main.tex
Paper: Meine Anmerkungen eingefuegt und teilweise Anmerkungen von CG bereits bearbeitet

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{abstract}
In this article, we describe the software library 
\textit{Differential Equations and Optimization Environment} (\dope{}).
The main feature of \dope{} is that it provides a unified interface to high level algorithms 
such as time-stepping methods, nonlinear solvers and optimization routines. This structure ensures 
that first of all the user is required to write only those sections of code that are specific to 
the considered problem. Second, the exchange of parts of the used routines is possible 
with only a few lines of code to change instead of large reimplementations.
The article illustrates the design principles and various features
of \dope{} and provides some 
numerical results as demonstration for the versatility of the software.
\end{abstract}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction}
\label{introduction}
Today, there exists a broad variety of software libraries
that provide numerical tools for solving partial differential
equations (PDEs). Prominent examples are 
\deal{} \citep{dealnew}, dune \citep{dune}, 
FEniCS \citep{fenics}
or commercial solvers as for instance ANSYS (with its Fluent package),
COMSOL, or simufact forming.
Finally, a lot of work is (still) done in Matlab.
In fact, these libraries provide very useful tools 
for a huge variety of applications.

@@

The project was initially initiated \todocg{initially initiated - doppelt gemoppelt?} at the University of Heidelberg in 2010 and  
at the present stage the following key features are supported by the library
\begin{itemize}
\item Solving of nonlinear stationary and nonstationary PDEs in 1d, 2d, and 3d.
\item Various, easily exchangeable, 
  time-stepping schemes (based on finite differences), 
  such as forward Euler, backward Euler,
  Crank-Nicolson, shifted Crank-Nicolson, and Fractional-Step-$\Theta$ scheme.
\item Use of all finite elements of from \deal{} including hp-support.
\item Some examples showing the solution of several PDEs including
   Poisson, Navier-Stokes, Plasticity and fluid-structure interaction problems. 
\item Line search and trust region newton algorithms for the 
   solution of optimization problems with PDE constraints by elimination of 
   the state variable.
\item Interface to SNOPT for the solution of optimization problems with PDEs and
  additional other control and state constraints.
\item Several examples showing how to solve various kinds of optimization 
  problems involving PDE constraints.
\item Goal-oriented mesh adaptation with the Dual Weighted Residual method and
  standard residual based error estimators.
\item Different spatial triangulations for control and state variables without
  the need to reimplement the user dependent code when switching from one common 
  triangulation to different ones.
\end{itemize}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Detailed Description of the Main Features}
\label{detailed_description}
This library is designed to allow easy implementation for the numerical solution 
of problems involving partial differential equations (PDEs)\todocg{Die Abkuerzung PDE wird auf Seite 1 schon erlaeuternt}. 
In particular, we remove all tedious routine implementation needed in 
\deal{} by introducing appropriate classes that take care of all the 
problem independent implementation while leaving only the problem specific 
parts to the user. To show how this is designed we will, in the sequel, 
describe the reasoning first for a simple stationary PDE in 
Section~\ref{subsubsec:stationary problems} and then show
how this idea extends to non stationary problems in 
Section~\ref{sec:timedep} as well as to optimization
problems with PDE constraints in Section~\ref{sec:opt}.

@@

In order to solve this problem, i.e., calculate a sequence of values $J(u_h)$ approximating $J(u)$,
we need to discretize 
the equation. In \dope, we are using finite elements to do this. This means we need to pick a 
subdivision $\mathcal T_h$, for instance into quadrilaterals,
of the domain $\Omega$ and a finite element. In the concrete case, we could, 
for instance, take the $\mathcal Q_2/\mathcal Q_1$ Taylor-Hood element.
As it is clear, that the initialization of the unknowns, the local shape 
functions, removal of hanging node constraints, etc. always follows the 
same procedure, we define an object which takes the finite element (FE)
and the triangulation and takes care of all other things related to
handling the finite element space $V_h$. We will refer, to 
this as the \texttt{SpaceTimeHandler}. Then, it is clear that the simplest 
form of the constructor should have the following interface:
\begin{lstlisting}
  template<typename TRIANGULATION, typename FE>
     void SpaceTimeHandler(TRIANGULATION& triag, FE& fe);
\end{lstlisting}
Next, one needs to solve the discrete equation
\begin{align}\label{eq:discrete_equation}
a(u_h)(\phi_h) = (f,\phi_h) \quad \forall \phi \in V_h.
\end{align}
To this end, one can employ a newton type method with some globalization 
like line-search. To do so, one needs to calculate the residual of the 
equation, as well as to solve linearized equations for the calculation of 
the newton method and then employing some kind of globalization strategy.
This again will usually require the evaluation of the residual.

@@

In the next section we will now sketch the way a user needs to implement the
user-dependent data as the local cell integrals and the initialization of the
objects described above.
\subsubsection{Problem dependent implementation of stationary PDE problems}
To solve a stationary PDE problem, the user has to provide some code which can roughly be divided into the following main sections:
\begin{description}
\item[\textit{Discretization}]\todocg{Hier scheint etwas verlorengegangen zu sein, der Satz macht so keinen Sinn:} Specify what kind of triangulation:
in most cases this is the
finite element and quadrature rules. Furthermore, other
things might be adjusted as well.
\item[\textit{Problem-data}] Define the weak form, the corresponding matrices, the
  functionals $J$ as well as the boundary conditions. With exception of the boundary conditions, we have to specify all this only cell-wise, \dope{} handles the rest for us.
\item[\textit{Solvers}] Choice of a (non-)linear solver. 
\item[\textit{Additional features}] ParameterReader to read in the run-time parameters
  such as material coefficients or adjustments for the solvers like tolerance
  for Newton's method. 
\end{description}

@@

\subsubsection{\texttt{localfunc.h}}
The implementation of the boundary-functional $J$ follows the outline of the
computation of the cell-residual in \texttt{localpde.h}. 
The main difference, and common to all functionals is that we only want to
evaluate a functional \todocg{Ist hier was verrutscht? der Satz passt im Moment nicht so wirklich, der Teil mit dem Vektor ist irgendwie verlorengegangen?}. Thus, we 
just loop over the quadrature points and not over the degrees of freedom 
on the face. In the present situation, we want to evaluate a functional 
that consists of a boundary integral. Thus we need to evaluate 
integrals on faces, those who lie on $\Gamma_{circ}$, instead of integrals over
elements. 
This face contributions are calculated in \texttt{BoundaryValue} and added to 
sum up to the value of $J$. 
\begin{lstlisting}
 double
 BoundaryValue(const FaceDataContainer& fdc)
  {
    //same as before, get n_q_points etc.
    const auto& face_values = cdc.GetFEFaceValuesState();
    fdc.GetFaceValuesState("state", ufacevalues);
    fdc.GetFaceGradsState("state", ufacegrads);
\end{lstlisting}
Notice that we get here the values of the computed solution by giving the string "state".
\begin{lstlisting}
    for(unsigned int q_point = 0; 
        q_point < n_q_points; 
        q_point++)
      erg -= ( _nu * ufacegrads[q_point][0]
               * face_values.normal_vector(q_point)
               + //other terms        
             ) * face_values.JxW(q_point);
    return erg;
  }
\end{lstlisting}

\subsection{Solving Nonstationary PDEs}\label{sec:timedep}
To describe the solution of nonstationary 
PDEs, we consider a time-dependent semi-linear form $a(u)(\phi)$ and some
right hand side functional $l$.\todocg{Warum hier l und nicht f?}
Let $I:=[0,T]$ be a time interval with end time point value $T$.
To find some $u$ in a weak form, we define
\[
\int_I a(u)(\phi)\, dt = \int_I l(\phi)\, dt \quad \forall\Phi\in X,
\]
with some suitable Bochner space $X$.  As model problem,
we extend the previous example and consider the nonstationary Navier-Stokes equations
in a domain $\Omega\in\mathbb{R}^d$. Find: 
$(v,p)\in X$ with
\begin{align*}
&\int_I \bigl[ (\partial_t v, \phi)
+ \nu (\nabla v, \nabla \phi) + (v\cdot\nabla v, \phi)
- (p,\nabla\cdot \phi)
+ (\nabla\cdot v, \chi)\bigr] \, dt\\
&\quad + (v(0) - v^0, \phi(0))
= \int_I l(\phi) \, dt 
\end{align*}
for all $\Phi:= (\phi, \chi) \in X$ with
\[
X:= \set{ 
(v,p) | v\in L^2(I,H_0^1(\Omega)^d) , 
\partial_t v\in L^2(I, L^2(\Omega)^d), 
p\in L^2(I,L^2(\Omega)\setminus\mathbb{R}}.
\]
As for stationary problems, the goal is to compute
some target functional $J$ which is now possibly 
time-dependent. To organize the discretization 
process, we need to perform spatial and temporal 
discretization. This is again done via
an object that takes care of organizing the degrees of 
freedom, as in the case of stationary problems.
The only difference to the stationary case is that we need to 
add the temporal subdivision \texttt{times}. In the simplest 
case, when we have the same spatial triangulation 
at all times, the constructor could look as follows:
\begin{lstlisting}
  template<typename TIMES, typename TRIANGULATION, 
           typename FE>
   void SpaceTimeHandler(TIMES& times,
                         TRIANGULATION& triag,
                         FE& fe);
\end{lstlisting}

@@

\section*{Thanks}
The \dope{} project makes use of various finite elements taken from  
the \deal{} \cite{dealnew} finite element library which has been developed
 initially by W. Bangerth, R. Hartmann, and G. Kanschat \cite{dealold}.
The authors acknowledge their past experience as well as discussions 
on modularization of algorithms
with 
the authors of the libraries 
Gascoigne/RoDoBo project, which was initiated by 
Roland Becker, Dominik Meidner, and Boris Vexler \cite{rodobo}. 
The second author thanks Mary F. Wheeler (ICES, Austin) for the 
possibility to finish this work.

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{abstract}
\todo{Wird die Webseite genannt? ist das sinnvoll dies zu tun?}
In this article, we describe the software library 
\textit{Differential Equations and Optimization Environment} (\dope{}).
The main feature of \dope{} is that it provides a unified interface to high level algorithms 
such as time-stepping methods, nonlinear solvers and optimization routines. This structure ensures 
that first of all the user is required to write only those sections of code that are specific to 
the considered problem. Second, the exchange of parts of the used routines is possible 
with only a few lines of code to change instead of large reimplementations.
The article illustrates the design principles and various features
of \dope{} and provides some 
numerical results as demonstration for the versatility of the software.
\end{abstract}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction}
\label{introduction}
Today, there exists a broad variety of software libraries
that provide numerical tools for solving partial differential
equations (PDEs). Prominent examples are 
\deal{} \citep{dealnew}, dune \citep{dune}, 
FEniCS \citep{fenics}
or commercial solvers as for instance ANSYS (with its Fluent package),
COMSOL, or simufact forming.
Finally, a lot of work is (still) done in Matlab.
In fact, these libraries provide very useful tools 
for a huge variety of applications.

@@

\todo{Das wir in Heidelberg angefangen haben wird schon vorher erwaehnt!
  Wuerde den Satz einfach streichen}
The project was initially initiated \todocg{initially initiated - doppelt gemoppelt?} at the University of Heidelberg in 2010 and  
at the present stage the following key features are supported by the library
\begin{itemize}
\item Solving of nonlinear stationary and nonstationary PDEs in 1d, 2d, and 3d.
\item Various, easily exchangeable, 
  time-stepping schemes (based on finite differences), 
  such as forward Euler, backward Euler,
  Crank-Nicolson, shifted Crank-Nicolson, and Fractional-Step-$\Theta$ scheme.
\item Use of all finite elements of from \deal{} including hp-support.
\item Some examples showing the solution of several PDEs including
   Poisson, Navier-Stokes, Plasticity and fluid-structure interaction problems. 
\item Line search and trust region Newton algorithms for the 
   solution of optimization problems with PDE constraints by elimination of 
   the state variable.
\item Interface to SNOPT for the solution of optimization problems with PDEs and
  additional other control and state constraints.
\item Several examples showing how to solve various kinds of optimization 
  problems involving PDE constraints.
\item Goal-oriented mesh adaptation with the Dual Weighted Residual method and
  standard residual based error estimators.
\item Different spatial triangulations for control and state variables without
  the need to reimplement the user dependent code when switching from one common 
  triangulation to different ones.
\end{itemize}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Detailed Description of the Main Features}
\label{detailed_description}
This library is designed to allow easy implementation for the numerical solution 
of problems involving PDEs.
In particular, we remove all tedious routine implementation needed in 
\deal{} by introducing appropriate classes that take care of all the 
problem independent implementation while leaving only the problem specific 
parts to the user. To show how this is designed we will, in the sequel, 
describe the reasoning first for a simple stationary PDE in 
Section~\ref{subsubsec:stationary problems} and then show
how this idea extends to non stationary problems in 
Section~\ref{sec:timedep} as well as to optimization
problems with PDE constraints in Section~\ref{sec:opt}.

@@

In order to solve this problem, i.e., calculate a sequence of values $J(u_h)$ approximating $J(u)$,
we need to discretize 
the equation. In \dope, we are using finite elements to do this. This means we need to pick a 
subdivision $\mathcal T_h$, for instance into quadrilaterals,
of the domain $\Omega$ and a finite element. In the concrete case, we could, 
\todo{Warum Tylor-Hood? Ist inf-sup stabil wuerde ich erwaehnen}
for instance, take the $\mathcal Q_2/\mathcal Q_1$ Taylor-Hood element.
As it is clear, that the initialization of the unknowns, the local shape 
functions, removal of hanging node constraints, etc. always follows the 
same procedure, we define an object which takes the finite element (FE)
and the triangulation and takes care of all other things related to
handling the finite element space $V_h$. We will refer, to 
this as the \texttt{SpaceTimeHandler}. Then, it is clear that the simplest 
form of the constructor should have the following interface:
\begin{lstlisting}
  template<typename TRIANGULATION, typename FE>
     void SpaceTimeHandler(TRIANGULATION& triag, FE& fe);
\end{lstlisting}
Next, one needs to solve the discrete equation
\begin{align}\label{eq:discrete_equation}
a(u_h)(\phi_h) = (f,\phi_h) \quad \forall \phi \in V_h.
\end{align}
\todo{wir schreiben newton immer klein. eigentlich ist das ein Eigenname: Newton}
To this end, one can employ a newton type method with some globalization 
like line-search. To do so, one needs to calculate the residual of the 
equation, as well as to solve linearized equations for the calculation of 
the newton method and then employing some kind of globalization strategy.
This again will usually require the evaluation of the residual.

@@

In the next section we will now sketch the way a user needs to implement the
user-dependent data as the local cell integrals and the initialization of the
objects described above.
\subsubsection{Problem dependent implementation of stationary PDE problems}
To solve a stationary PDE problem, the user has to provide some code which can roughly be divided into the following main sections:
\begin{description}
\item[\textit{Discretization}]\todocg{Hier scheint etwas verlorengegangen zu
    sein, der Satz macht so keinen Sinn:} 
\todo{Vorschlag: Specification of triangulation issues such as finite elements
  and quadrature rules}  
Specify what kind of triangulation:
in most cases this is the
finite element and quadrature rules. Furthermore, other
things might be adjusted as well.
\item[\textit{Problem-data}] \todo{Definition of } Define the weak form, the corresponding matrices, the
  functionals $J$ as well as the boundary conditions. With exception of the boundary conditions, we have to specify all this only cell-wise, \dope{} handles the rest for us.
\item[\textit{Solvers}] Choice of a (non-)linear solver. 
\item[\textit{Additional features}] ParameterReader to read in the run-time parameters
  such as material coefficients or adjustments for the solvers like tolerance
  for Newton's method. 
\end{description}

@@

\todo{Ich wuerde localfunctional als volles Wort schreiben}
\subsubsection{\texttt{localfunc.h}}
The implementation of the boundary-functional $J$ follows the outline of the
computation of the cell-residual in \texttt{localpde.h}. 
The main difference, and common to all functionals is that we only want to
evaluate a functional \todocg{Ist hier was verrutscht? der Satz passt im
  Moment nicht so wirklich, der Teil mit dem Vektor ist irgendwie
  verlorengegangen?}. 
\todo{Ich hatte Deine Anmerkung so verstanden, als ob Du alles nach dem i.e.,
  geloescht 
haben wolltest. Im main TW steht noch das Original}
Thus, we 
just loop over the quadrature points and not over the degrees of freedom 
on the face. In the present situation, we want to evaluate a functional 
that consists of a boundary integral. Thus we need to evaluate 
integrals on faces, those who lie on $\Gamma_{circ}$, instead of integrals over
elements. 
This face contributions are calculated in \texttt{BoundaryValue} and added to 
sum up to the value of $J$. 
\begin{lstlisting}
 double
 BoundaryValue(const FaceDataContainer& fdc)
  {
    //same as before, get n_q_points etc.
    const auto& face_values = cdc.GetFEFaceValuesState();
    fdc.GetFaceValuesState("state", ufacevalues);
    fdc.GetFaceGradsState("state", ufacegrads);
\end{lstlisting}
Notice that we get here the values of the computed solution by giving the string "state".
\begin{lstlisting}
    for(unsigned int q_point = 0; 
        q_point < n_q_points; 
        q_point++)
      erg -= ( _nu * ufacegrads[q_point][0]
               * face_values.normal_vector(q_point)
               + //other terms        
             ) * face_values.JxW(q_point);
    return erg;
  }
\end{lstlisting}

\subsection{Solving Nonstationary PDEs}\label{sec:timedep}
To describe the solution of nonstationary 
PDEs, we consider a time-dependent semi-linear form $a(u)(\phi)$ and some
right hand side functional $l$.\todocg{Warum hier l und nicht f?}
\todo{Antwort: Weil $f$ die volumen kraft ist. Wir wollten 
doch Funktional und Kraftvektor unterscheiden. Daher hier $l$. Es gilt :
$l(\phi) = (f,\phi)$.}
Let $I:=[0,T]$ be a time interval with end time point value $T$.
To find some $u$ in a weak form, we define
\[
\int_I a(u)(\phi)\, dt = \int_I l(\phi)\, dt \quad \forall\Phi\in X,
\]
with some suitable Bochner space $X$.  As model problem,
we extend the previous example and consider the nonstationary Navier-Stokes equations
in a domain $\Omega\in\mathbb{R}^d$. Find: 
$(v,p)\in X$ with
\begin{align*}
&\int_I \bigl[ (\partial_t v, \phi)
+ \nu (\nabla v, \nabla \phi) + (v\cdot\nabla v, \phi)
- (p,\nabla\cdot \phi)
+ (\nabla\cdot v, \chi)\bigr] \, dt\\
&\quad + (v(0) - v^0, \phi(0))
= \int_I l(\phi) \, dt 
\end{align*}
for all $\Phi:= (\phi, \chi) \in X$ with
\[
X:= \set{ 
(v,p) | v\in L^2(I,H_0^1(\Omega)^d) , 
\partial_t v\in L^2(I, L^2(\Omega)^d), 
p\in L^2(I,L^2(\Omega)\setminus\mathbb{R}}.
\]
As for stationary problems, the goal is to compute
some target functional $J$ which is now possibly 
time-dependent. To organize the discretization 
process, we need to perform spatial and temporal 
discretization. This is again done via
an object that takes care of organizing the degrees of 
freedom, as in the case of stationary problems.
The only difference to the stationary case is that we need to 
add the temporal subdivision \texttt{times}. In the simplest 
case, when we have the same spatial triangulation 
at all times, the constructor could look as follows:
\begin{lstlisting}
  template<typename TIMES, typename TRIANGULATION, 
           typename FE>
   void SpaceTimeHandler(TIMES& times,
                         TRIANGULATION& triag,
                         FE& fe);
\end{lstlisting}

@@

\section*{Thanks}
The \dope{} project makes use of various finite elements taken from  
the \deal{} \cite{dealnew} finite element library which has been developed
 initially by W. Bangerth, R. Hartmann, and G. Kanschat \cite{dealold}.
The authors acknowledge their past experience as well as discussions 
on modularization of algorithms
with 
the authors of the libraries 
Gascoigne/RoDoBo project, which was initiated by 
Roland Becker, Dominik Meidner, and Boris Vexler \cite{rodobo}. 
Finally, the second author thanks Mary F. Wheeler (ICES, Austin) for the 
possibility to finish this work.
